Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-07-19 ms_pre_post.Rmd
Address Reviewer 1 comments about model restrictions

Inbreeding depression is widely hypothesised to drive adaptive evolution of pre-copulatory and post-copulatory mechanisms of inbreeding avoidance, which in turn are hypothesised to affect evolution of polyandry (i.e., female multiple mating). However, surprisingly little theory or modelling critically examines selection for pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, given evolutionary constraints and direct costs, or examines how evolution of inbreeding avoidance strategies might feed back to affect evolution of polyandry. Selection for post-copulatory inbreeding avoidance, but not for pre-copulatory inbreeding avoidance, requires polyandry, while interactions between pre-copulatory and post-copulatory inbreeding avoidance might cause functional redundancy (i.e., 'degeneracy') potentially generating complex evolutionary dynamics among inbreeding strategies and polyandry. We use individual-based modelling to quantify evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression and different evolutionary constraints and direct costs. We find that evolution of post-copulatory inbreeding avoidance increases selection for initially rare polyandry, and that selection for a costly inbreeding avoidance strategy was weak or negligible given a lower cost alternative strategy. Further, fixed pre-copulatory inbreeding avoidance often completely precluded evolution of polyandry and hence post-copulatory inbreeding avoidance, but fixed post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance. Evolution of inbreeding avoidance phenotypes and associated polyandry are therefore affected by evolutionary feedbacks and degeneracy. All else being equal, evolution of pre-copulatory inbreeding avoidance and resulting low polyandry is more likely when post-copulatory inbreeding avoidance is precluded or costly, and evolution of post-copulatory inbreeding avoidance greatly facilitates evolution of costly polyandry.

@@

Overall, polyandry can therefore simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry has consequently been widely hypothesised to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002]. Resulting indirect selection on polyandry that stems from the increase in offspring fitness could explain evolution of polyandry in cases where multiple mating does not increase female reproductive success, or even reduces reproductive success and therefore imposes direct costs on polyandrous females. However, despite such widely-invoked hypotheses regarding the role of inbreeding depression in driving pre-copulatory and post-copulatory strategies and polyandry, there is surprisingly little theory or modelling that critically examines the conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are predicted to evolve, or that examines how the evolution of such strategies of inbreeding avoidance might feed back to affect underlying evolution of polyandry.

Despite the paucity of theory, numerous empirical studies on diverse species have tested for, and in some cases found evidence of, female inbreeding avoidance in systems where polyandry also occurs [@Tregenza2002; @Varian-Ramos2012; @Kingma2013; @Arct2015; but see @Reid2015a]. However, relatively few studies have determined whether inbreeding avoidance is enacted through pre-copulatory or post-copulatory mechanisms. Pre-copulatory inbreeding avoidance has been reported in diverse species, spanning sweet potato weevils [*Cylas formicarius*, @Kuriwada2011], purple-crowned fairy-wrens [*Malurus coronatus*, @Kuriwada2011], and squinting bush brown butterflies [*Bicyclus anynana*, @Fischer2015], while post-copulatory inbreeding avoidance has been reported in, **e.g.,** red junglefowl [*Gallus gallus*, @Pizzari2004] and crickets [*Teleogryllus oceanicus*, @Simmons2006; *Gryllus bimaculatus*, @Bretman2009]. Evidence for both pre-copulatory and post-copulatory inbreeding avoidance is available across different studies of Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. Meanwhile, @Liu2014 found evidence of pre-copulatory but not post-copulatory inbreeding avoidance within a single study of cabbage beetles (*Colaphellus bowringi*). However, @Ala-Honkola2011 and @Tan2012 found no evidence for pre-copulatory or post-copulatory inbreeding avoidance in fruit flies (*Drosophila melanogaster*), respectively, and Reid et al. [-@Reid2014; -@Reid2015] found no net inbreeding avoidance in song sparrows (*Melospiza melodia*) despite strong inbreeding depression and opportunities for both pre-copulatory and post-copulatory inbreeding avoidance. Taken together, these studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance, or lack thereof, occur in nature. However, there is as yet no theory that predicts what combinations of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry should be favoured by selection when all can evolve. Consequently, there is no theory that allows the diversity of observed pre-copulatory and post-copulatory inbreeding avoidance strategies to be interpreted, and no clear hypotheses that could be tested through future empirical studies of individual systems or subsequent comparative analyses.

@@

The few previous models that considered the evolution of inbreeding avoidance through mate choice (as opposed to dispersal) have implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models are useful for isolating the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect the realised degree of inbreeding, it cannot be assumed that both strategies will simultaneously invade a randomly mating population, nor that the invasion fitness of one strategy will be independent of the pre-existence or invasion fitness of the other strategy. For example, if pre-adaptation or a selective sweep results in fixation of alleles underlying pre-copulatory inbreeding avoidance, then new alleles underlying polyandry and post-copulatory inbreeding avoidance might be unlikely to invade a population because the phenotypic effect of such alleles on the overall degree of inbreeding, and resulting indirect selection, could be negligible. Conversely, fixation of alleles underlying polyandry and post-copulatory inbreeding avoidance might reduce or eliminate positive selection on alleles underlying pre-copulatory inbreeding avoidance and hence impede adaptive evolution of mate choice. New theory, guided by modelling that evaluates invasion dynamics of alleles underlying multiple interacting and potentially functionally redundant (i.e., `degenerate') traits, is therefore needed.

@@

We use individual-based modelling to address three key questions regarding evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given opportunity for inbreeding and strong inbreeding depression. First, does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained by increasing relatedness of polyandrous females' additional mates? Second, how do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of these reproductive strategies given cost asymmetry? Third, how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? **For each question, we isolate the effect of the biological mechanism of interest (i.e., post-copulatory inbreeding avoidance, cost asymmetry, and another fixed strategy of inbreeding avoidance) by running simulations in which the mechanism is present and comparing the results to those of simulations in which the it is absent but all other parameter values are identical** By addressing these **three** questions, we illustrate how the simultaneous evolution of multiple interacting degenerate reproductive strategies can generate diverse evolutionary outcomes.

@@

In all simulations, we recorded mean values of $P_{a}$, $M_{a}$, and $F_{a}$ in each generation and present these values over generations to infer selection on phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$). Each combination of parameter values simulated was replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allowed us to infer how allele values changed over generations in response to costs, but also in response to the changing values of other alleles and therefore potential evolutionary feedbacks among reproductive strategies. For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be sufficient for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling showed that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but not so small that selection is typically overwhelmed by drift [@Duthie2016a].

>>>>

Inbreeding depression is widely hypothesised to drive adaptive evolution of pre-copulatory and post-copulatory mechanisms of inbreeding avoidance, which in turn are hypothesised to affect evolution of polyandry (i.e., female multiple mating). However, surprisingly little theory or modelling critically examines selection for pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, given evolutionary constraints and direct costs, or examines how evolution of inbreeding avoidance strategies might feed back to affect evolution of polyandry. Selection for post-copulatory inbreeding avoidance, but not for pre-copulatory inbreeding avoidance, requires polyandry, while interactions between pre-copulatory and post-copulatory inbreeding avoidance might cause functional redundancy (i.e., 'degeneracy') potentially generating complex evolutionary dynamics among inbreeding strategies and polyandry. We use individual-based modelling to quantify evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression and different evolutionary constraints and direct costs. We find that evolution of post-copulatory inbreeding avoidance increases selection for initially rare polyandry, and that evolution of a costly inbreeding avoidance strategy became negligible over time given a lower cost alternative strategy. Further, fixed pre-copulatory inbreeding avoidance often completely precluded evolution of polyandry and hence post-copulatory inbreeding avoidance, but fixed post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance. Evolution of inbreeding avoidance phenotypes and associated polyandry are therefore affected by evolutionary feedbacks and degeneracy. All else being equal, evolution of pre-copulatory inbreeding avoidance and resulting low polyandry is more likely when post-copulatory inbreeding avoidance is precluded or costly, and evolution of post-copulatory inbreeding avoidance greatly facilitates evolution of costly polyandry.

@@

Overall, polyandry can therefore simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry has consequently been widely hypothesised to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002]. Resulting indirect selection on polyandry that stems from the increase in offspring fitness could explain evolution of polyandry in cases where multiple mating does not increase female reproductive success, or even reduces reproductive success and therefore imposes direct costs on polyandrous females. However, despite such widely-invoked hypotheses regarding the role of inbreeding depression in driving pre-copulatory and post-copulatory strategies and polyandry, there is surprisingly little theory or modelling that critically examines the conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are predicted to evolve, or that examines how the evolution of such strategies of inbreeding avoidance might feed back to affect underlying evolution of polyandry. **To comprehensively understand evolution of mating strategies given inbreeding depression, it is necessary to consider the fundamental effects of selection on pre-copulatory and post-copulatory inbreeding avoidance and polyandry.**

Despite the paucity of theory, numerous empirical studies on diverse species have tested for, and in some cases found evidence of, female inbreeding avoidance in systems where polyandry also occurs [@Tregenza2002; @Varian-Ramos2012; @Kingma2013; @Arct2015; but see @Reid2015a]. However, few studies have determined whether inbreeding avoidance is enacted through pre-copulatory or post-copulatory mechanisms. Pre-copulatory inbreeding avoidance has been reported in sweet potato weevils [*Cylas formicarius*, @Kuriwada2011], purple-crowned fairy-wrens [*Malurus coronatus*, @Kuriwada2011], and squinting bush brown butterflies [*Bicyclus anynana*, @Fischer2015], while post-copulatory inbreeding avoidance has been reported in, **e.g.,** red junglefowl [*Gallus gallus*, @Pizzari2004] and crickets [*Teleogryllus oceanicus*, @Simmons2006; *Gryllus bimaculatus*, @Bretman2009]. Evidence for both pre-copulatory and post-copulatory inbreeding avoidance is available across different studies of Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. Meanwhile, @Liu2014 found evidence of pre-copulatory but not post-copulatory inbreeding avoidance within a single study of cabbage beetles (*Colaphellus bowringi*). However, @Ala-Honkola2011 and @Tan2012 found no evidence for pre-copulatory or post-copulatory inbreeding avoidance in fruit flies (*Drosophila melanogaster*), respectively, and Reid et al. [-@Reid2014; -@Reid2015] found no net inbreeding avoidance in song sparrows (*Melospiza melodia*) despite strong inbreeding depression and opportunities for both pre-copulatory and post-copulatory inbreeding avoidance. Taken together, these studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance, or lack thereof, occur in nature. However, there is as yet no theory that predicts what combinations of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry should be favoured by selection when all can evolve. Consequently, there is no theory that allows the diversity of observed pre-copulatory and post-copulatory inbreeding avoidance strategies to be interpreted, and no clear hypotheses that could be tested through future empirical studies of individual systems or subsequent comparative analyses.

@@

The few previous models that considered evolution of inbreeding avoidance through mate choice (as opposed to dispersal) have implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models are useful for isolating the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect the realised degree of inbreeding, it cannot be assumed that both strategies will simultaneously invade a randomly mating population, nor that the invasion fitness of one strategy will be independent of the pre-existence or invasion fitness of the other strategy. For example, if pre-adaptation or a selective sweep results in fixation of alleles underlying pre-copulatory inbreeding avoidance, then new alleles underlying polyandry and post-copulatory inbreeding avoidance might be unlikely to invade a population because the phenotypic effect of such alleles on the overall degree of inbreeding, and resulting indirect selection, could be negligible. Conversely, fixation of alleles underlying polyandry and post-copulatory inbreeding avoidance might reduce or eliminate positive selection on alleles underlying pre-copulatory inbreeding avoidance and hence impede adaptive evolution of mate choice. New theory, guided by modelling that evaluates invasion dynamics of alleles underlying multiple interacting and potentially functionally redundant (i.e., `degenerate') traits, is therefore needed.

@@

We use individual-based modelling to address three key questions regarding evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given opportunity for inbreeding and strong inbreeding depression. First, does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained by increasing relatedness of polyandrous females' additional mates? Second, how do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of these reproductive strategies given cost asymmetry? Third, how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? **For each question, we isolate the effect of the biological mechanism of interest (i.e., post-copulatory inbreeding avoidance, cost asymmetry, and another fixed strategy of inbreeding avoidance) by running simulations in which the mechanism is present and comparing the results to those of simulations in which the it is absent but all other parameter values are identical.** **We restrict our model to simulate how evolution of traits are affected by their direct costs and indirect benefits due to inbreeding avoidance. Although empirical populations will include additional beneficial traits that might affect evolution of mating strategies, targeting our focus on traits affecting inbreeding and polyandry allows us to better isolate their effects and therefore better develop hypotheses for empirical systems.** By addressing **our three** questions, we illustrate how the simultaneous evolution of multiple interacting degenerate reproductive strategies can generate diverse evolutionary outcomes.

@@

In all simulations, we recorded mean values of $P_{a}$, $M_{a}$, and $F_{a}$ in each generation and present these values over generations to infer selection on phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$). Each combination of parameter values simulated was replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allowed us to infer how allele values changed over generations in response to costs, but also in response to the changing values of other alleles and therefore potential evolutionary feedbacks among reproductive strategies. **In our analyses, we do not use statistical tests to interpret simulation results; such tests are innappropriate because they violate the underlying assumptions of statistical hypothesis testing, and because statistical power (and therefore p-values) would be entirely determined by the number of simulation replicates performed [@White2014].** For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be sufficient for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling showed that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but not so small that selection is typically overwhelmed by drift [@Duthie2016a].
